### 2025-01-28 nondirectional_ms_2024.Rmd
zip new result files for all analyses + Finalizing the Rmd + minor edit in appendix H + updated run of appenidx H for SC + add absolute effect size test to all analysis + update references + make appendix D more empirically based + remove differences plot from appenidx E + separate appedix H to abses and sc + calculte phi effect size for categorical data in Battich + re-run all analyses

authornote: |
  All data except for the datasets from Machado et al., 2007, 2009, for which participants did not consent to having their data shared online, and all analysis code are available at https://github.com/mufcItay/NDT. This study's analysis was not pre-registered.

  All authors declare no conflict of interests.

abstract: |
  A wave of criticisms and replication failures is currently challenging claims about the scope of unconscious perception and cognition. Such failures to find unconscious processing effects at the population level may reflect the absence of individual-level effects, or alternatively, the averaging out of individual-level effects with opposing signs. Importantly, only the first suggests that consciousness may be necessary for the tested process to take place. To arbitrate between these two possibilities, we tested previously collected data where unconscious processing effects were not found (26 effects from 470 participants), using four frequentist and Bayesian tests that are robust to individual differences in effect signs. By and large, we found no reliable evidence for unconscious effects being masked by individual differences. In contrast, when we examined 135 non-significant effects from other domains, a novel non-parametric sign consistency test did reveal effects that were hidden by opposing individual results, though as we show, some of them might be driven by design-related factors. Taken together, four analysis approaches provide strong evidence for the restricted nature of unconscious processing effects not only across participants, but also across different trials within individuals. We provide analysis code and best-practice recommendations for testing for non-directional effects.

@@

library(groundhog)
groundhog.library(c(
  'ggplot2',
  'stringr',
  'gridExtra',
  'ggh4x',
  'foreach',
  'ggtext',
  'ggridges',
  'ggpubr',
  'scales',
  'dplyr',
  'tidyr',
  'pracma',
  'extraDistr',
  'patchwork',
  'MCMCpack',
  'BayesFactor',
  'xfun'), '2023-07-22')
# from github (Commit: 0e04c3c)
library(signcon)

@@

# sourcing the 'fig1sim' script generates the panels populating Figure 1 in the manuscript, and analyzes the simulated data with all available non-directional (here we used an edited version of this panels)
# tests, and t-tests.
source('figures\\fig1sim.R')
# sourcing the 'fig2_3' script generates figure 2, the lower panel of figure 3, and Appendix C,G (note that figure 3 was edited to also include the illustration in the upper panel of figure 3)
# it also analyzes the results of all unconscious processing datasets according to all tests
source('figures\\fig2_3.R')
# sourcing the 'fig4' script generates figure 4 (minor edits were applied to the resulting svg figure).
# it also analyzes the results of all datasets (including datasets from non-UC domains) according to the sign consistency test.
source('figures\\fig4.R')

@@

# the analysis and simulations for appendix D.
# to run the analysis again, uncomment the (commented) run_appendixD line and comment the read.csv line which reads the results of an existing simulations run.
source('appendix\\appendix_D.R')
# appendixD_res <- run_appendixD(apndxD_conf_FAs, apndxD_conf_sensitivity)
appendixD_res <- read.csv('appendix\\results\\2024-01-07_23_39_16Appendix_D_QUID_OANOVA.csv')

# the analysis and simulations for appendix E.
# to run the analysis again, uncomment the (commented) run_appendixE line and comment the read.csv line which reads the results of an existing simulations run (keeping the save_plot_appendixE line.)
source('appendix\\appendix_E.R')
# appendixE_res <- run_appendixE(apndx_e_conf)
appendixE_res <- read.csv('appendix\\results\\2023-12-29_18_48_11Appendix_E.csv')
save_plot_appendixE(appendixE_res)

@@

# the analysis and simulations for appendix H.
# to run the analysis again, uncomment the (commented) run_appendixH line and comment the read.csv line which reads the results of an existing simulations run.
source('appendix\\appendix_H.R')
apndxH_N_p <- N_p
apndxH_N_t <- N_t
# appendixH_res <- run_appendixH(apndx_e_conf)
appendixH_res <- read.csv('appendix\\results\\2024-01-09_16_46_09Appendix_H.csv')
apndx_H_median_cnd_power <- appendixH_res %>%
  filter(N_p == median(appendixH_res$N_p), 
         N_t == median(appendixH_res$N_t),
         sds_ratio == median(appendixH_res$sds_ratio)) %>%
  pull(Power)

@@

For example, Vorberg and colleagues [-@vorberg2003a] studied the role of consciousness in motor preparation. In a series of experiments, they presented an arrow stimulus (henceforth, the prime) which was followed by an arrow-shaped metacontrast target stimulus, rendering it invisible (@breitmeyer1976a; see Figure 1A). Unconscious motor preparation priming was demonstrated by showing that participants responded faster to the target stimulus when its direction was congruent with the direction of the prime. This suggests that the direction of the prime has been represented unconsciously, triggering a motor plan. In similar studies, participants were reported to unconsciously perform high-level functions such as arithmetic operations [@sklar2012a], extract and integrate word meanings [@damian2001a; @gaal2014a; @sklar2012a], or scenes and objects [@mudrik2011a], detect errors [@charles2013a], and exert inhibition over responses [@gaal2008a] to stimuli that were masked from awareness. Findings of high-level processing in the absence of consciousness served to inform and reform theories of consciousness [@dehaene2001a; @oizumi2014a; @lamme2020a; @lau2011a].

@@

On the face of it, pronounced individual differences in unconscious processing effects on cognition and perception seem possible, even likely. Indeed, previous investigations revealed heterogeneity in susceptibility to the attentional blink [@martens2006a], in the effects of stimulus onset asynchrony (SOA) on metacontrast masking [@albrecht2010a], the speed of breaking perceptual suppression [@sklar2021a], both in the breaking continuous flash suppression paradigm (b-CFS; [@jiang2007a] and the breaking repeated mask suppression one (b-RMS; [@abir2020a], and in the effects of visual imagery on conscious perception in a binocular rivalry setting [@dijkstra2019a]. Some qualitative differences have been linked to variability in processing speed [@martens2006a], genetics [@maksimov2013a], and brain anatomy and physiology [@gaal2011a; @boy2010a]. Critically, in other behavioural paradigms, unconscious stimuli had opposite effects on different participants. Bolger and colleagues [-@bolger2019a] showed that while most participants responded faster to upright faces in a b-CFS task, some responded faster to upside-down faces. Other findings from several different study groups have repeatedly revealed that masked priming effects changed in magnitude and even flipped in sign as a function of the interval between prime and target [@boy2014a; @boy2010b; @parkinson2014a; @schlaghecken2004a], or the presentation duration of stimuli rendered unconscious using either crowding [@faivre2011a] or CFS [@faivre2012a; @barbot2012a]. Moreover, different hypotheses were laid out over the years regarding the driving mechanisms of the counter-intuitive 'negative' priming effects. Among others, response inhibition of initial prime activations [@eimer2003a; @boy2010c], or neural habituation [@faivre2011a; @faivre2012a; @barbot2012a; @jacob2021a] were suggested. Taken together, it is not clear if, and to what extent, unconscious effects are subject to meaningful individual variability. Crucially, if they are, then some previously reported null results might actually be true effects, masked by such variability.

The paper proceeds as follows. We first simulate a setting where a strong effect of unconscious processing on behaviour is entirely missed in standard analysis, due to pronounced inter-individual differences. We then show that the same effect is revealed when using three tests that are robust to population variability: the global null prevalence test [@donhauser2018a], Bayesian hierarchical modelling [@haaf2019a], and a test based on analysis of variance (ANOVA; [@miller2018a]. Importantly, unlike common measures of reliability which are used to directly estimate individual differences (see @parsons2019a), the above tests do not quantify individual differences, but *measure group effects in a way that is robust to such differences*. Hence, they provide researchers with the appropriate tools for detecting unconscious effects even if pronounced individual differences exist, without depending on that being the case.

We then apply these tests to data gathered from eight unconscious processing studies (reporting 26 non-significant effects), and show that the same three tests support the null hypothesis according to which the behaviour of individual participants is unaffected by unconscious cognition and perception. This strengthens claims for a true absence of an effect in these studies. Finally, we propose a non-parametric alternative that provides improved sensitivity and specificity, avoiding potentially unjustified statistical assumptions regarding the data-generating process. Our test successfully reveals effects on multisensory integration, visual search, and confidence ratings that could not be detected using standard directional analysis. However, similar to the three other approaches, it reveals no effects when applied to the studies of unconscious processing examined here. We conclude that existing data are most consistent with the absence of influences of unconscious stimuli on cognition and perception, not only at the population, but also at the single-participant level.

@@

In two simulations, we manipulated two factors: the between-participant standard deviation (SD) over effect sizes ($\sigma_{b}$), and the within-participant SD over RTs within each condition ($\sigma_{w}$). This resulted in two distinct scenarios under this framework: (1) a *qualitative* or *non-directional differences* scenario, where all individuals show an effect, but individual-level effects largely vary in magnitude and sign ($\sigma_{b}$=`r nde_sd_b`, $\sigma_{w}$=`r nde_sd_w`; Figure 1B)), and (2) a *global null* scenario [@allefeld2016a; @nichols2005a], where no single participant is affected by the experimental manipulation ($\sigma_{b}$=`r sn_sd_b`, $\sigma_{w}$= `r sn_sd_w`; Figure 1C). We simulated $N_{t}$=`r 2*sn_trials` trials from $N_{p}$=`r sn_n` participants per scenario, noting that the general principle holds for other sample sizes and number of trials.

@@

To reiterate, a standard t-test misses existing individual-level effects because, operating on individual-level summary statistics, it is oblivious to within-participant variability in the dependent variable. In recent years, researchers sought to address this limitation, advocating for the use of statistical methods that incorporate both within and between-participant variability. Specifically, three approaches were proposed. First, the *prevalence global null* approach (@donhauser2018a; henceforth GNT) tests if the prevalence of individual-level effects in a given population (the proportion of individuals showing an effect) is greater than zero. The prevalence approach relies on a two stages procedure. In the first stage, effects are tested at the individual level using a standard hypothesis-testing approach. In the second stage, the proportion of observed individual-level effects is tested against the alpha level, or type-1 error rate, using a binomial test. A significantly higher prevalence means the *global null* hypothesis, according to which no individual shows a true effect, can be rejected. Second, the *qualitative individual differences* approach (@rouder2020a; @rouder2021a; @haaf2019a; henceforth QUID) quantifies the relative support for the presence of “qualitative differences” in effects, that is, inter-individual differences in effect signs, by performing a Bayesian model comparison over a family of hierarchical models with different constraints [@haaf2019a]. Third, Miller & Schwarz [-@miller2018a] introduce a parametric and frequentist test, based on ANOVA. Specifically, their Omnibus ANOVA test (henceforth OANOVA) probes the joint null hypothesis that there are no systematic differences neither between experimental conditions across individuals, nor within individuals and across trials. Together, this is equivalent to the *global null* scenario we presented above.

We applied the tests to our simulated data. For QUID, we used the default priors from the original publication [@rouder2021a]. For GNT and OANOVA, we used an $\alpha$ of `r oanova_alpha` to examine individual-level and group-level effects. For QUID, we considered $BF>`r bf_criteria`$ as evidence for an effect, $BF< \frac{1}{`r bf_criteria`}$ as evidence for no effect (*global null*), and values between these thresholds ($\frac{1}{`r bf_criteria`}\leq BF\leq `r bf_criteria`$) as inconclusive [@jeffreys1998a]. Reassuringly, all tests were able to differentiate between the two simulated scenarios, providing very strong evidence for an effect in the *non-directional differences* scenario, but not in the *global null* one. Specifically, According to GNT, the prevalence of effects on RT was clearly above zero in the *non-directional differences* simulation (using a two-sided t-test for the individual-level test; `r round(100 * nde_gnt_res$stat,2)`% of significant effects, one-sided $CI_{95}$ = [`r round(100 * nde_gnt_res$ci_low,2)` `r round(100 * nde_gnt_res$ci_high,2)`], p `r apa_p(nde_gnt_res$p)`), but this proportion was not higher than $\alpha$ in the *global null* simulation (`r round(100 * sn_gnt_res$stat,2)`% of significant effects, one-sided $CI_{95}$ = [`r round(100 * sn_gnt_res$ci_low,2)` `r round(100 * sn_gnt_res$ci_high,2)`], p =`r apa_p(sn_gnt_res$p)`). Using the QUID method, a random effects model with individual-level effects was overwhelmingly preferred in the *non-directional differences* simulation ($BF$= `r formatC(nde_bf, format = 'e', 2)`), but a null model was preferred in the *global null* simulation ($BF$= `r round(sn_bf,2)`). Similarly, the OANOVA test revealed significant results in the *non-directional differences* scenario (F(`r nde_OANOVA_res$df_num`, `r nde_OANOVA_res$df_denom`) = `r round(nde_OANOVA_res$F,2)`, p `r paste0(ifelse(nde_OANOVA_res$p >= .001, '=', ''), apa_p(nde_OANOVA_res$p))`), and a non-significant effect in the *global null* simulation (F(`r sn_OANOVA_res$df_num`, `r sn_OANOVA_res$df_denom`) = `r round(sn_OANOVA_res$F,2)`, p `r paste0(ifelse(sn_OANOVA_res$p >= .001, '=', ''),apa_p(sn_OANOVA_res$p))`).

@@

(ref:figure1caption) *Note*. Simulated data demonstrating how true effects of unconscious priming can be masked by heterogeneity at the population level. Panel A: stimuli in a typical unconscious processing experiment (based on Vorberg and colleagues, [-@vorberg2003a]). Participants make speeded decisions about a consciously perceived target stimulus (for example, the direction of an arrow: right or left). The presentation of the target stimulus is preceded by a prime stimulus, which is masked from awareness. Decision time is measured as a function of prime-target agreement: congruent (blue) or incongruent (red). Panels B, C: Left: simulation parameters controlling the within ($\sigma_w$) and between ($\sigma_b$) participant SD. Right: the results that were generated using the simulation parameters. Each point depicts the measured individual-level summary statistics for the difference between the mean RTs of each condition (congruent and incongruent), and the blue and red segments depict the 95% confidence interval (CI) around the average of RTs (the grey segment in the middle of each CI) in the congruent and incongruent conditions, respectively. A constant of `r offset_rt`ms was added to the RTs in both panels for presentation purposes. Panel B: a *non-directional differences* scenario (simulated using the parameters $\sigma_{b}$=`r nde_sd_b`, $\sigma_{w}$=`r nde_sd_w`). Panel C: a *global null* scenario (no effect of the experimental manipulation; simulated using the parameters $\sigma_{b}$=`r sn_sd_b`, $\sigma_{w}$=`r sn_sd_w`). Since standard directional tests rely on individual-level summary statistics, they cannot arbitrate between the scenarios described in the two panels.
```{r figure1, out.width="100%", fig.cap = "(ref:figure1caption)"} 
knitr::include_graphics("figures//figure1.png")
```

# Reexamining unconscious effects
```{r empiric-all-tests, include=FALSE}
# get the counts of experiments for the section (uc datasets), reading from the enviroment created by soucing the fig2_3.R script
n_empiric_ns <- uc_ns_results %>% filter(is_sim == FALSE) %>% pull(exp) %>% unique() %>% length()
sc_sig_n <- uc_ns_results %>% filter(is_sim == FALSE, uc_ns_results$signcon.p < .05) %>% pull(exp) %>% unique() %>% length()
n_empiric_ns_quid <- ns_quid_res %>% filter(is_sim == FALSE) %>% pull(exp) %>% unique() %>% length()
n_empiric_ns_oanova <- ns_OANOVA_res %>% filter(is_sim == FALSE) %>% pull(exp) %>% unique() %>% length()
p_gnt_stat_zero <- ns_gnt_res %>% filter(is_sim == FALSE, gnt.stat == 0)%>% pull(exp) %>% unique() %>% length() / n_empiric_ns
p_null_QUID <- ns_quid_res %>% filter(is_sim == FALSE, quid_bf < 1/bf_criteria)%>% pull(exp) %>% unique() %>% length() / n_empiric_ns_quid
# get the results of individual experiments with significant sign consistency.
# in the results data frame there are multiple lines per exp (listing the null distribution per experiment). We average across rows to get the repeating statistics per experiment.
res_per_exp <- uc_ns_results %>% group_by(exp) %>% summarise_all(mean)
skora_1_res <- res_per_exp %>% filter(exp == 'S1')
skora_1_corrected_p <- all_datasets_signcon$nsdir_res %>% filter(exp == "Skora et al_2020_1") %>% pull(test_result_corrected)
bm_3_res <- res_per_exp %>% filter(exp == 'BM3')
bm_3_corrected_p <- all_datasets_signcon$nsdir_res %>% filter(exp == "Biderman & Mudrik_2018_3") %>% pull(test_result_corrected)

>>>>

authornote: |
  All data except for the datasets from Machado et al., 2007, 2009, for which participants did not consent to having their data shared online, and all analysis code are available at https://github.com/mufcItay/NDT. This study's analysis was not pre-registered, and all authors declare no conflict of interests.

abstract: |
  A wave of criticisms and replication failures is currently challenging claims about the scope of unconscious perception and cognition. Such failures to find unconscious processing effects at the population level may reflect the absence of individual-level effects, or alternatively, the averaging out of individual-level effects with opposing signs. Importantly, only the first suggests that consciousness may be necessary for the tested process to take place. To arbitrate between these two possibilities, we tested previously collected data where unconscious processing effects were not found (26 effects from 470 participants), using five frequentist and Bayesian tests that are robust to individual differences in effect signs. By and large, we found no reliable evidence for unconscious effects being masked by individual differences. In contrast, when we examined 135 non-significant effects from other domains, two novel non-parametric tests did reveal effects that were hidden by opposing individual results, though as we show, some of them might be driven by design-related factors. Taken together, five analysis approaches provide strong evidence for the restricted nature of unconscious processing effects not only across participants, but also across different trials within individuals. We provide analysis code and best-practice recommendations for testing for non-directional effects.

@@

library(groundhog)
groundhog.library(c(
  'ggplot2',
  'stringr',
  'gridExtra',
  'ggh4x',
  'foreach',
  'ggtext',
  'ggridges',
  'ggpubr',
  'scales',
  'dplyr',
  'tidyr',
  'pracma',
  'extraDistr',
  'patchwork',
  'MCMCpack',
  'BayesFactor',
  'xfun'), '2023-07-22')
# from github (Commit: 9d338dd)
library(signcon)

@@

# sourcing the 'fig1sim' script generates the panels populating Figure 1 in the manuscript, and analyzes the simulated data with all available non-directional (here we used an edited version of this panels)
# tests, and t-tests.
source('figures\\fig1sim.R')
# sourcing the 'fig2_3' script generates figure 2, the lower panel of figure 3, and Appendix C,G (note that figure 3 was edited to also include the illustration in the upper panel of figure 3)
# it also analyzes the results of all unconscious processing datasets according to all tests
source('figures\\fig2_3.R')
# sourcing the 'fig4' script generates figure 4 (minor edits were applied to the resulting svg figure).
# it also analyzes the results of all datasets (including datasets from non-UC domains) according to the sign consistency and absolute effect size tests.
source('figures\\fig4.R')

@@

# the analysis and simulations for appendix D.
# to run the analysis again, uncomment the (commented) run_appendixD line and comment the read.csv line which reads the results of an existing simulations run.
source('appendix\\appendix_D.R')
# appendixD_res <- run_appendixD(apndxD_conf_FAs, apndxD_conf_sensitivity)
appendixD_res <- read.csv('appendix\\results\\2025-01-20_21_09_19Appendix_D_QUID_OANOVA.csv')

# the analysis and simulations for appendix E.
# to run the analysis again, uncomment the (commented) run_appendixE line and comment the read.csv line which reads the results of an existing simulations run (keeping the save_plot_appendixE line.)
source('appendix\\appendix_E.R')
# appendixE_res <- run_appendixE(apndx_e_conf)
appendixE_res <- read.csv('appendix\\results\\2025-01-26_07_53_42Appendix_E.csv')
save_plot_appendixE(appendixE_res)

@@

# the analysis and simulations for appendix H.
# to run the analysis again, uncomment the (commented) run_appendixH line and comment the read.csv line which reads the results of an existing simulations run.
source('appendix\\appendix_H.R')
apndxH_N_p <- N_p
apndxH_N_t <- N_t
# appendixH_res <- run_appendixH(apndx_e_conf)
appendixH_res <- read.csv('appendix\\results\\2025-01-26_17_09_45Appendix_H_ ABSES.csv')
apndx_H_median_cnd_power_abs_es <- appendixH_res %>%
  filter(N_p == median(appendixH_res$N_p), 
         N_t == median(appendixH_res$N_t),
         sds_ratio == median(appendixH_res$sds_ratio)) %>%
  pull(Power)

@@

For example, Dehaene and colleagues [-@dehaene1998a] studied the role of consciousness in semantic processing. In their seminal study, they presented a number word stimulus (henceforth, the prime), followed and preceded by strings of random letters, acting as backward and forward masks, rendering it invisible. Then, a fully visible number stimulus was presented (henceforth, the target; see Figure 1A). Participants were instructed to report whether the target stimulus was greater or smaller than the number five. Unconscious semantic priming was demonstrated by showing that participants responded faster in congruent trials, when the target and the prime  were both smaller or larger than five (see again Figure 1A). This suggests that the numerical magnitude of the prime was processed unconsciously, affecting the response to the visible target number (see @damian2001a and @naccache2001a for critical assessment and discussions). In similar studies, participants were reported to unconsciously perform other high-level functions such as arithmetic operations [@sklar2012a], extract and integrate word meanings [@damian2001a; @gaal2014a; @sklar2012a], or scenes and objects [@mudrik2011a], detect errors [@charles2013a], and exert inhibition over responses [@gaal2008a] to stimuli that were masked from awareness. Findings of high-level processing in the absence of consciousness served to inform and reform theories of consciousness [@dehaene2001a; @oizumi2014a; @lamme2020a; @lau2011a].

@@

On the face of it, pronounced individual differences in unconscious processing effects on cognition and perception seem possible, even likely. Indeed, using objective measures of awareness such as discrimination ability (direct tasks; @schmidt2006a and @schmidt2024a, previous research revealed  heterogeneity in participants’ susceptibility to different masking paradigms, and in the effects of design choices on participants’ ability to consciously perceive the masked stimulus. Participants have been shown to reliably vary in their susceptibility to the attentional blink [@martens2006a], and in the speed of breaking perceptual suppression— both in the breaking continuous flash suppression paradigm (b-CFS; [@sklar2021a] and the breaking repeated mask suppression one (b-RMS; [@abir2020a]. Furthermore,  manipulating stimulus onset asynchrony (SOA, @albrecht2010a; see also @biafora2020a) and mask contrast (@biafora2020a) has been shown to produce different, sometimes opposite effects on metacontrast masking in different participants. Finally, visual imagery had different effects on the conscious perception of different participants in a binocular rivalry setting [@dijkstra2019a]. Some qualitative differences have been linked to variability in processing speed [@martens2006a], genetics [@maksimov2013a], and brain anatomy and physiology [@gaal2011a; @boy2010a].

Critically, unconscious processing effects (i.e. indirect tasks; @schmidt2006a) have also been shown to vary as a function of task parameters and across participants. At the heart of these findings are reports of “negative” priming effects: cases where the processing of the target was facilitated by an incongruent prime, rather than a congruent one. Indeed, masked priming effects changed in magnitude and even flipped in sign as a function of the interval between prime and target [@boy2014a; @boy2010b; @parkinson2014a; @schlaghecken2004a], or the presentation duration of stimuli rendered unconscious using either crowding [@faivre2011a] or CFS [@faivre2012a; @barbot2012a]. Negative effects were also observed for higher level features such as perceptual expectations: Bolger and colleagues [-@bolger2019a] showed that while most participants responded faster to upright faces in a b-CFS task, some responded faster to upside-down faces. Different hypotheses were laid out over the years regarding the driving mechanisms of the counter-intuitive negative priming effects. Among others, response inhibition of initial prime activations [@eimer2003a; @boy2010c], or neural habituation [@faivre2011a; @faivre2012a; @barbot2012a; @jacob2021a] were suggested. Taken together, it is not clear if, and to what extent, unconscious effects are subject to meaningful individual variability. Crucially, if they are, then some previously reported null results might actually be true effects, masked by such variability.

The paper proceeds as follows. We first simulate a setting where a strong effect of unconscious processing on behaviour is entirely missed in standard analysis, due to pronounced inter-individual differences. We then show that the same effect is revealed when using three tests that are robust to population variability: the global null prevalence test [@donhauser2018a], Bayesian hierarchical modelling [@haaf2019a], and a test based on analysis of variance (ANOVA; [@miller2018a]. Importantly, unlike common measures of reliability which are used to directly estimate individual differences (see @parsons2019a), the above tests do not quantify individual differences, but measure group effects in a way that is robust to such differences. Hence, they provide researchers with the appropriate tools for detecting unconscious effects even if pronounced individual differences exist, without depending on that being the case.

We then apply these tests to data gathered from eight unconscious processing studies (reporting 26 non-significant effects), and show that the same three tests support the null hypothesis according to which the behaviour of individual participants is unaffected by unconscious cognition and perception. This strengthens claims for a true absence of an effect in these studies. Finally, we propose two non-parametric alternatives that provide improved sensitivity and specificity, avoiding potentially unjustified statistical assumptions regarding the data-generating process. Our tests successfully reveal effects on multisensory integration, visual search, spatial attention and confidence ratings that could not be detected using standard directional analysis. However, similar to the three other approaches, our tests reveal no effects when applied to the studies of unconscious processing examined here. We conclude that existing data are most consistent with the absence of influences of unconscious stimuli on cognition and perception, not only at the population, but also at the single-participant level.

@@

In two simulations, we manipulated two factors: the between-participant standard deviation (SD) over effect sizes ($\sigma_{b}$), and the within-participant SD over RTs within each condition ($\sigma_{w}$). This resulted in two distinct scenarios under this framework: (1) a *qualitative* or *non-directional differences* scenario, where the parameters of the generative model for all individuals were set to generate a true, non-zero effect (i.e., $e_i \neq 0$,  for all individuals), but individual-level effects largely vary in magnitude and sign ($\sigma_{b}$=`r nde_sd_b`, $\sigma_{w}$=`r nde_sd_w`; Figure 1B), and (2) a *global null* scenario [@allefeld2016a; @nichols2005a], where no single participant is affected by the experimental manipulation ($\sigma_{b}$=`r sn_sd_b`, $\sigma_{w}$= `r sn_sd_w`; Figure 1C). We simulated $N_{t}$=`r sn_trials` trials per condition from $N_{p}$=`r sn_n` participants per scenario, noting that the general principle holds for other sample sizes and number of trials.

@@

To reiterate, a standard t-test misses existing individual-level effects because, operating on individual-level summary statistics, it cannot differentiate between within-participant variability in the dependent variable (noise) and meaningful between-participant variability. In recent years, researchers sought to address this limitation, advocating for the use of *non-directional* statistical methods that incorporate both within and between-participant variability. Here, we examined three non-directional approaches, all capable of detecting effects that exist within at least a single participant, without relying on group-level assumptions about the consistency of effect signs across individuals. As a result, a non-directional finding indicates that an effect exists within single participants without committing to whether the average effect at the group-level is positive, negative, or null.

First, the *prevalence global null* approach (@donhauser2018a; henceforth GNT) tests if the prevalence of individual-level effects in a given population (the proportion of individuals showing an effect) is greater than zero. The prevalence approach relies on a two-stage procedure. In the first stage, effects are tested at the individual level using a standard hypothesis-testing approach for a given significance level ($\alpha_{individual}$). In the second stage, the proportion of observed individual-level effects (the observed *prevalence*) is tested against $\alpha_{individual}$, which is the type-1 error rate of the individual-level test. This is done, using a one-sided binomial test (for which a separate significance level is used, here termed $\alpha_{prevalence}$, which may be different from $\alpha_{individual}$). Due to the one-sidedness of GNT, all confidence intervals on the prevalence of effects include 100% by definition (i.e., all individuals show an effect), and the crucial aspect is whether the lower bound of the confidence interval exceeds $\alpha_{individual}$. Hence, a significantly higher prevalence than $\alpha_{individual}$ means that the *global null* hypothesis, according to which no individual shows a true effect, can be rejected.

Second, the *qualitative individual differences* approach (@rouder2021a; @haaf2019a; henceforth QUID) quantifies the relative support for the presence of “qualitative differences” in effects, that is, inter-individual differences in effect signs, by performing a Bayesian model comparison over a family of hierarchical models with different constraints [@haaf2019a]. Specifically, QUID models reaction times (RTs) using a standard linear model, where individual-level effects are treated as random effects. Then, to quantify the support for an effect while allowing for qualitative individual differences, it relies on Bayesian Factors (BFs). The BF analysis compares evidence of a model that poses no constraints over the magnitude of individual-level effects, with a model in which all participants show no effect (see @rouder2021a for further model comparisons, aimed at examining different hypotheses regarding the character of population variability in effects).

Third, Miller & Schwarz [-@miller2018a] introduce a parametric and frequentist test, based on ANOVA. Specifically, their Omnibus ANOVA test (henceforth OANOVA) probes the joint null hypothesis that there are no systematic differences neither between experimental conditions across individuals, nor within individuals and across trials. Together, this is equivalent to the *global null* scenario we presented above. OANOVA relies on a trial-level ANOVA model, comparing the variability in the dependent variable which is explained by the combination of the experimental manipulation and its interaction with the participants against the variability explained solely by participants' intercepts. Hence, OANOVA considers participants as fixed effects in the analysis, and enables detecting effects without assuming homogeneity in effect signs.

We applied the tests to our simulated data. For QUID, we used the default priors from the original publication [@rouder2021a]. For GNT and OANOVA, we used an $\alpha$ of `r oanova_alpha` to examine individual-level and group-level effects. For QUID, we considered $BF>`r bf_criteria`$ as evidence for an effect, $BF< \frac{1}{`r bf_criteria`}$ as evidence for no effect (*global null*), and values between these thresholds ($\frac{1}{`r bf_criteria`}\leq BF\leq `r bf_criteria`$) as inconclusive [@jeffreys1998a]. Reassuringly, all tests were able to differentiate between the two simulated scenarios, providing very strong evidence for an effect in the *non-directional differences* scenario, but not in the *global null* one. Specifically, the prevalence of effects on RT was clearly above the expected proportion of significant effects ($\alpha_{individual}$)  in the *non-directional differences* simulation (using a two-sided t-test for the individual-level test; `r round(100 * nde_gnt_res$stat)`% significant effects, one-sided $CI_{95}$ = [`r round(100 * nde_gnt_res$ci_low)`, `r round(100 * nde_gnt_res$ci_high)`], p `r apa_p(nde_gnt_res$p)`), but this proportion was not higher than $\alpha_{individual}$ in the *global null* simulation (`r round(100 * sn_gnt_res$stat)`% significant effects, one-sided $CI_{95}$ = [`r round(100 * sn_gnt_res$ci_low)`, `r round(100 * sn_gnt_res$ci_high)`], p =`r apa_p(sn_gnt_res$p)`). Using the QUID method, a random effects model with individual-level effects was overwhelmingly preferred in the *non-directional differences* simulation ($BF$= `r formatC(nde_bf, format = 'e', 2)`), but a null model was preferred in the *global null* simulation ($BF$= `r round(sn_bf,2)`). Similarly, the OANOVA test revealed significant results in the *non-directional differences* scenario (F(`r nde_OANOVA_res$df_num`, `r nde_OANOVA_res$df_denom`) = `r round(nde_OANOVA_res$F,2)`, p `r paste0(ifelse(nde_OANOVA_res$p >= .001, '=', ''), apa_p(nde_OANOVA_res$p))`), and a non-significant effect in the *global null* simulation (F(`r sn_OANOVA_res$df_num`, `r sn_OANOVA_res$df_denom`) = `r round(sn_OANOVA_res$F,2)`, p `r paste0(ifelse(sn_OANOVA_res$p >= .001, '=', ''),apa_p(sn_OANOVA_res$p))`).

@@

(ref:figure1caption) *Note*. Simulated data demonstrating how true effects of unconscious priming can be masked by heterogeneity at the population level. Panel A: stimuli in a typical unconscious processing experiment (based on Dehaene and colleagues, [-@dehaene1998a]). Participants make speeded decisions about a consciously perceived target stimulus (for example, its magnitude: being larger or smaller than the number five). The presentation of the target stimulus is preceded by a prime stimulus, which is masked from awareness. Decision time is measured as a function of prime-target agreement: congruent (blue) or incongruent (red). Panels B, C: Left: simulation parameters controlling the within ($\sigma_w$) and between ($\sigma_b$) participant SD. Right: the results that were generated using the simulation parameters. Each point depicts the measured individual-level summary statistics for the difference between the mean RTs of each condition (congruent and incongruent) with 95% confidence interval ($CI_{95}$) around each difference estimate, and the blue and red segments depict the $CI_{95}$ around the average of RTs in each condition separately (the grey segment in the middle of each CI) in the congruent and incongruent conditions, respectively. A constant of `r offset_rt`ms was added to the RTs in both panels for presentation purposes. Panel B: a *non-directional differences* scenario (simulated using the parameters $\sigma_{b}$=`r nde_sd_b`, $\sigma_{w}$=`r nde_sd_w`). Panel C: a *global null* scenario (no effect of the experimental manipulation; simulated using the parameters $\sigma_{b}$=`r sn_sd_b`, $\sigma_{w}$=`r sn_sd_w`). Since standard directional tests rely on individual-level summary statistics, they cannot arbitrate between the scenarios described in the two panels.
```{r figure1, out.width="100%", fig.cap = "(ref:figure1caption)"} 
knitr::include_graphics("figures//figure1.png")
```

# Reexamining unconscious effects
```{r empiric-all-tests, include=FALSE}
# get the counts of experiments for the section (uc datasets), reading from the enviroment created by soucing the fig2_3.R script
n_empiric_ns <- uc_ns_results %>% filter(is_sim == FALSE) %>% pull(exp) %>% unique() %>% length()
sc_sig_n <- uc_ns_results %>% filter(is_sim == FALSE, uc_ns_results$signcon.p < .05) %>% pull(exp) %>% unique() %>% length()
abs_es_sig_n <- uc_ns_results %>% filter(is_sim == FALSE, uc_ns_results$absolute_es.p < .05) %>% pull(exp) %>% unique() %>% length()
n_empiric_ns_quid <- ns_quid_res %>% filter(is_sim == FALSE) %>% pull(exp) %>% unique() %>% length()
n_empiric_ns_oanova <- ns_OANOVA_res %>% filter(is_sim == FALSE) %>% pull(exp) %>% unique() %>% length()
p_gnt_stat_zero <- ns_gnt_res %>% filter(is_sim == FALSE, gnt.stat == 0)%>% pull(exp) %>% unique() %>% length() / n_empiric_ns
p_null_QUID <- ns_quid_res %>% filter(is_sim == FALSE, quid_bf < 1/bf_criteria)%>% pull(exp) %>% unique() %>% length() / n_empiric_ns_quid
# get the results of individual experiments with significant sign consistency and absolute effect size.
# in the results data frame there are multiple lines per exp (listing the null distribution per experiment). We average across rows to get the repeating statistics per experiment.
res_per_exp <- uc_ns_results %>% group_by(exp) %>% summarise_all(mean)
skora_1_res <- res_per_exp %>% filter(exp == 'S1')
skora_2_res <- res_per_exp %>% filter(exp == 'S2')
bm_3_res <- res_per_exp %>% filter(exp == 'BM3')
# significant sign consistency correction:
skora_1_corrected_p_sc <- all_datasets_signcon$nsdir_res %>% filter(exp == "Skora et al_2020_1") %>% pull(test_result_corrected)
bm_3_corrected_p_sc <- all_datasets_signcon$nsdir_res %>% filter(exp == "Biderman & Mudrik_2018_3") %>% pull(test_result_corrected)
# significant absolute effect size correction:
skora_1_corrected_p_abs_es <- all_datasets_absES$nsdir_res %>% filter(exp == "Skora et al_2020_1") %>% pull(test_result_corrected)
skora_2_corrected_p_abs_es <- all_datasets_absES$nsdir_res %>% filter(exp == "Skora et al_2020_2") %>% pull(test_result_corrected)
